### 2019-02-18 sections/intuition.tex
cigarette smokers

The step combinator is one form of internal communication, where task values are passed to continuations.
Another form of internal communication is shared data.
Shared data enables data flow \emph{across} control flow, in particular between parallel tasks.
Shared data sources are assignable references whose changes are immediately visible to all tasks interested in them.
The semantics of \TOPHAT requires all updates to shared data and all enabled internal steps to be processed before any further communication with the environment can take place.

@@

The cigarette smokers problem \cite{books/Downey08LBOS} is a surprisingly tricky synchronisation problem.
We study it here because it demonstrates the capabilities of guarded steps.
% The problem requires waiting for two conditions, waking up only if both conditions are satisfied.
The problem is stated as follows.
In order to smoke a cigarette, three ingredients are required: tobacco, paper, and a match.
There are three smokers, each having one of the ingredients and requiring the other two.
There is an agent that randomly provides two of the ingredients.

Downey models availability of the ingredients using a semaphore for each ingredient.
The agent randomly signals two of the three semaphores.
% The problem asks writing code for the smokers so that the smoker who requires the two signalled semaphores wakes up and proceeds with smoking.
%
% A naïve attempt where each smoker first waits for one and then the other semaphore can lead to deadlock.
% The following pseudo-code is a non-solution.
% Let \emph{paper}, \emph{tobacco} and \emph{match} be semaphores, initialised to 0.
% \begin{TASK}
%   tobaccoSmoker = wait(match); wait(paper); smoke();
%   paperSmoker = wait(tobacco); wait(match); smoke();
%   matchSmoker = wait(tobacco); wait(paper); smoke();
% \end{TASK}
% To see what can go wrong, imagine the three smokers running in parallel.
% Let the agent signal the two semaphores \emph{tobacco} and \emph{paper}.
% One of \emph{paperSmoker} and \emph{matchSmoker} can take their first step, but if \emph{paperSmoker} takes the step, the system deadlocks.
%
The solution proposed by Downey involves an additional mutex, three additional semaphores, three additional threads called \emph{pushers}, and three regular Boolean variables.
The job of the pushers is to record availability of their ingredient in their Boolean variable, and check availability of other resources, waking the correct smoker when appropriate.

The details of the solution are not important here.
What is important is that the implementation of what is essentially waiting for two semaphores requires a substantial amount of additional synchronisation, together with non-trivial conditional statements.

% A similar effort as in Downey's solution must be made in order to solve the problem using synchronised actions in \CSP.
% This is because in \CSP, processes can only synchronise on one action at a time.

\TOPHAT allows a simple solution to this problem, using guarded steps.
Steps can be guarded with arbitrary expressions, and the parallel combinator can be used to watch two shared editors ($\Update$) at the same time.
Let \TS{match}, \TS{paper}, and \TS{tobacco} be references to Booleans.
The smokers are defined as follows.
\begin{TASK}
  let continue = \<<x,y>>. if x /\ y then smoke else fail in
  let tobaccoSmoker = (update match <&> update paper) >>? continue in
  let paperSmoker = (update tobacco <&> update match) >>? continue in
  let matchSmoker = (update tobacco <&> update paper) >>? continue
\end{TASK}
When the agent supplies two of the ingredients by setting the respective shares to \TS{True}, only the step of the smoker that waits for those becomes enabled.

>>>>

The step combinator is one form of internal communication, where task values are passed to continuations.
Another form of internal communication is shared data.
Shared data enables data flow \emph{across} control flow, in particular between parallel tasks.
Shared data sources are assignable references whose changes are immediately visible to all tasks interested in them.
If $r$ is a reference of type $\tau$, then $\Update r$ is an editor whose value is that of $r$.

The semantics of \TOPHAT requires all updates to shared data and all enabled internal steps to be processed before any further communication with the environment can take place.

@@

The cigarette smokers problem \cite{books/Downey08LBOS} is a surprisingly tricky synchronisation problem.
We study it here because it demonstrates the capabilities of guarded steps.
% The problem requires waiting for two conditions, waking up only if both conditions are satisfied.
The problem is stated as follows.
In order to smoke a cigarette, three ingredients are required: tobacco, paper, and a match.
There are three smokers, each having one of the ingredients and requiring the other two.
There is an agent that randomly provides two of the ingredients.
The difficulty lies in the requirement that only the smoker may proceed whose missing ingredients are present.

Downey models availability of the ingredients with a semaphore for each ingredient.
The agent randomly signals two of the three semaphores.
The solution proposed by Downey involves an additional mutex, three additional semaphores, three additional threads called \emph{pushers}, and three regular Boolean variables.
The job of the pushers is to record availability of their ingredient in their Boolean variable, and check availability of other resources, waking the correct smoker when appropriate.

The details of the solution are not important here.
What is important is that the implementation of what is essentially deadlock-free waiting for two semaphores requires a substantial amount of additional synchronisation, together with non-trivial conditional statements.

\TOPHAT allows a simple solution to this problem, using guarded steps.
Steps can be guarded with arbitrary expressions, and the parallel combinator can be used to watch two shared editors at the same time.
Let \TS{match}, \TS{paper}, and \TS{tobacco} be references to Booleans.
The smokers are defined as follows.
\begin{TASK}
  let continue = \<<x,y>>. if x /\ y then smoke else fail in
  let tobaccoSmoker = (update match <&> update paper) >>? continue in
  let paperSmoker = (update tobacco <&> update match) >>? continue in
  let matchSmoker = (update tobacco <&> update paper) >>? continue
\end{TASK}
When the agent supplies two of the ingredients by setting the respective shares to \TS{True}, only the step of the smoker that waits for those becomes enabled. \qed
